Predicting Fire Risk: \\ An Evaluation of the Fire Danger Index

### 2016-01-29 paper.tex
added figures
plotting code
Updated latex

\usepackage{fullpage,booktabs,amsmath}
\usepackage[section]{placeins}
\usepackage{gensymb,microtype,todonotes}
\usepackage{natbib}
\usepackage{pdfpages}
\usepackage{listings}

@@

McArthur's FFDI originated as tabulated indices and were subsequently reduced to equations by \citet{noble80} which in turn simplified into the formulation of the FFDI given by
\begin{equation}
  \label{eq FFDI}
  F=2.0 e^{-0.450 + 0.987 \ln(D)-0.0345H+0.0338T+0.0234V},
\end{equation}
where
\begin{itemize}
  \item $D$ is the drought factor index;
  \item $H$ is relative humidity (as a percentage);
  \item $T$ is temperature (degrees Celsius); and
  \item $V$ is average wind velocity (in km/hr).
\end{itemize}

@@

Evaluations of other fire danger indices and weather related indices around the world have used statistical techniques as simple as correlation, and as sophisticated as logistic regression (\cite{andrews03}; \cite{harris14}).

@@

\begin{figure}
  \centering
  %	\includegraphics[width=1\columnwidth]{figures/Data/cfadis.pdf}
  \caption{Country Fire Authority Districts}
  \label{fig:dist}
\end{figure}

@@

%Graphs, need to redo better
\begin{figure}
  \centering
  %	\includegraphics[width=1\columnwidth]{figures/Data/silvan.pdf}
  \caption{Time series of meteorological variables for Silvan (-37.82, 145.42)}
  \label{fig:silv}
\end{figure}

@@

Combination of these variables as per equation (\ref{eq FFDI}) yields the FFDI which clearly exhibits seasonality. Note the spike in summer 2009, corresponding to the Black Saturday bushfires.

%WS, Temp and FDR contours, see if can change colours to match with the diagram
\begin{figure}
  \centering
  %	\includegraphics[width=1\columnwidth]{figures/Data/tw_fdr.pdf}
  \caption{Wind Speed and temperature coloured by fire danger rating}
  \label{fig:tw_fdr}
\end{figure}

The FFDI is most sensitive to temperature and windspeed (\cite{dowdy10}). Figure~\ref{fig:tw_fdr} shows two-dimensional density plots for windspeed and temperature coloured by the fire danger rating (only observations from October--March included) and divided by CFA district. The relative sizes of each contour show the small proportion of extreme and code red days and the difference in weather conditions observed in different parts of Victoria. For instance the only districts observing code red days are districts 16, 17 and 18.

@@

%\includepdf[pages={1}]{figures/Data/curdist.pdf}

% level plot of GS curing, probably needs resizing
\begin{figure}
  \centering
  %	\includegraphics[width=1\columnwidth]{figures/Data/cur_lp.pdf}
  \caption{Level Plot of Grassland Curing }
  \label{fig:cur_lp}
\end{figure}

@@

% level plot of vapour pressure, level plot may be unnesc could just have std plot
\begin{figure}
  \centering
  %	\includegraphics[width=1\columnwidth]{figures/Data/vap_ras.pdf}
  \caption{Raster plot of vapour pressure 9am }
  \label{fig:vap_ras}
\end{figure}

@@

%Commented out until final printing,
%Map of fire ignitions in Victoria
\begin{figure}
  \centering
  %	\includegraphics[width=1\columnwidth]{figures/data/firemap.pdf}
  \caption{Map of fire ignition 2000 - 2009}
  \label{fig:firemap}
\end{figure}

%timeseries histogram with count of ignitions per fortnight
\begin{figure}
  \centering
  %	\includegraphics[width=1\columnwidth]{figures/data/occurence.pdf}
  \caption{Number of ignitions by fortnight}
  \label{fig:occ}
\end{figure}

@@

\begin{figure}
  \centering
  %	\includegraphics[width=1\columnwidth]{figures/Data/tw_fire.pdf}
  \caption{Wind Speed and temperature coloured by fire/no fire}
  \label{fig:tw_fire}
\end{figure}

@@

\begin{figure}
  \centering
  %	\includegraphics[width=1\columnwidth]{figures/Data/dh_fire.pdf}
  \caption{Drought factor and relative humidity coloured by fire/no fire}
  \label{fig:dh_fire}
\end{figure}

@@

\begin{figure}
  \centering
  %	\includegraphics[width=1\columnwidth]{figures/Data/gv_fire.pdf}
  \caption{Grassland curing and vapour pressure coloured by fire/no fire}
  \label{fig:gv_fire}
\end{figure}

@@

%eg confusion matrix with ABCD
\begin{table}
  \centering
  \begin{tabular}{lrrrr}
    & 	\multicolumn{1}{l}{} & \multicolumn{2}{c}{\emph{Model Prediction}} \\
    & 	\multicolumn{1}{l}{} & \multicolumn{1}{r}{No Fire} & \multicolumn{1}{r}{Fire} \\
    \cline{3-5}
    \emph{Actual} & No Fire & $a$   & $b$   & $a+b$ \\
                  &   Fire  & $c$   & $d$   & $c + d$ \\
    \cline{3-5}
                  &         & $a+c$ & $b+d$ & $a+b+c+d$ \\
    \cline{3-5}
  \end{tabular}
  \caption{Example confusion matrix}
  \label{table:ex}
\end{table}
An example confusion matrix is given in Table~\ref{table:ex}.
From this matrix, we can compute accuracy, sensitivity (correctly classified fires dividied by total number of actual fires), specificity (correctly identified non-fires divided by total number of non-fires), balanced accuracy and kappa:
\begin{align}
  \label{eq acc}
  \mbox{Accuracy}          & = \frac{a + d}{a+b+c+d}                     \\
  \label{eq sens}
  \mbox{Sensitivity}       & = \frac{a }{a+b}                            \\
  \label{eq spec}
  \mbox{Specificity}       & = \frac{d}{c+d}                             \\
  \label{eq BA}
  \mbox{Balanced Accuracy} & = \frac{\frac{a }{a+b} + \frac{d}{c+d} }{2} \\
  \label{eq:K}
  \kappa                   & = \frac{P_0 - P_c}{1-P_c}
\end{align}
where $P_0$ is the total agreement probability and $P_c$ is the Agreement probability which is due to chance (see \cite{kappa} for definitions).

On top of the accuracy measures in equations (\ref{eq sens}), (\ref{eq spec}) and (\ref{eq BA}) the value in the bottom left of the confusion matrix (c) where we incorrectly predict `No fire' is very important in this context.

@@

I add splines to Model 3 in stages using the confusion matrices for 10-fold cross validation to assess accuracy (using the caret package by \cite{caret}). The best performing model based on balanced accuracy and number of false negatives (incorrectly predicted `No Fires') is reported in Table~\ref{table:logf}.

@@

Figure~\ref{fig:gam_pl} shows the spline functions for each of the inputs. In logistic models the spline functions are not the marginal effects of the variables on probability of ignition but the marginal effect on the log-odds ratio. Equation~\ref{eq logits} can be rearranged to give the log-odds ratio \citep{james13},
\begin{equation}
  \label{eq:logodds}
  \log(\frac{p(X)}{1-p(X)}) = \sum_{k=1}^{K} f_k (x_k).
\end{equation}

@@

%ROC for logits
\begin{figure}
  \centering
  %	\includegraphics[width=1\columnwidth]{figures/all_roc.pdf}
  \caption{ROC curve for logistic models}
  \label{fig:ffdi}
\end{figure}

@@

As classification trees can be sensitive to the sample used, I constructed several trees on different balanced samples of the data. As each of the trees were structured very similarly I only report two in figure~\ref{fig:tree2}.

% Trees with all variables
\begin{figure}
  \centering
  %	\includegraphics[page=1, width=1\columnwidth]{figures/trees2.pdf}
  %	\includegraphics[page=2, width=1\columnwidth]{figures/trees2.pdf}
  \caption{Trees with all variables}
  \label{fig:tree2}
\end{figure}

@@

Interpreting the first tree in figure~\ref{fig:tree2}, when the grassland curing level is outside the 39 to 97 range the tree predicts there will be no fire. Approximately 33\% of the training data lies in these branches.

@@

The classification tree results reaffirm the results of the spline models in figure~\ref{fig:gam_pl} which suggested the effect of curing and drought factor on probability of ignition is not constant at each level of the variable. The bands of values chosen by the tree as having higher probability of ignition vaguely match the shape of the fitted spline functions in Model 4. The tree selecting ranges of variables for which probability of ignition changes suggests there is a  non-linear relationship between some of the explanatory variables and probability of ignition.

@@

\begin{figure}
  \centering
  %	\includegraphics[width=1\columnwidth]{figures/trees_roc.pdf}
  \caption{ROC curve for Tree Models}
  \label{fig:troc}
\end{figure}

@@

The variable importance is calculated as mean decrease in accuracy using the out-of-bag observations, described in detail by \cite{archer08}. Figure~\ref{fig:vimp} shows the variable importance for the forest.

@@

Similar variable importance may indicate that on the average across all trees the variables are equally important in determining probability of ignition, however variable importance cannot necessarily be taken simply at face value. \cite{strob07}  and \cite{strob08} show that variable importance measures may be biased by the measurement scale of the variables and show a preference for correlated variables. Alternative calculation of variable importance is outside the scope of this paper but the limitations of the Gini method should be taken into account.

@@

\begin{figure}
  \centering
  %	\includegraphics[width=1\columnwidth]{figures/varimp.pdf}
  \caption{Variable Importance for random forest}
  \label{fig:vimp}
\end{figure}

Finally figure~\ref{fig:rocrf} shows the ROC curve for the random forest, a single classification tree, the linear logistic model and the splines model. Table~\ref{table:AUC} contains the area under the curves (AUC) for the ROC curves shown in~\ref{fig:rocrf} . We see the forest and splines model outperform a single tree and linear-logistic model. The closeness of the two models is interesting given random forests are generally expected to be a the best predictors at the expense of interpretability.

@@

\begin{figure}
  \centering
  %	\includegraphics[width=1\columnwidth]{figures/roc_rf.pdf}
  \caption{Variable Importance for random forest}
  \label{fig:rocrf}
\end{figure}

>>>>

\usepackage{fullpage,booktabs,amsmath,paralist}
\usepackage[section]{placeins}
\usepackage{gensymb,microtype,todonotes}
\usepackage{natbib}
\usepackage{pdfpages}
\usepackage{listings}

@@

McArthur's FFDI originated as tabulated indices and were subsequently reduced to equations by \citet{noble80} which in turn simplified into the formulation of the FFDI given by
\begin{equation}
  \label{eq FFDI}
  F=2.0 e^{-0.450 + 0.987 \ln(D)-0.0345H+0.0338T+0.0234V},
\end{equation}
where
\begin{compactitem}
  \item $D$ is the drought factor index;
  \item $H$ is relative humidity (as a percentage);
  \item $T$ is temperature (degrees Celsius); and
  \item $V$ is average wind velocity (in km/hr).
\end{compactitem}

@@

Evaluations of other fire danger indices and weather related indices around the world have used statistical techniques as simple as correlation, and as sophisticated as logistic regression \citep{andrews03,harris14}.

@@

\begin{figure}
  \centering
  \includegraphics[width=1\columnwidth]{figures/cfadis.pdf}
  \caption{Country Fire Authority Districts}
  \label{fig:dist}
\end{figure}

\todo[inline]{Change Figure \ref{fig:dist} to have statistical divisions}

@@

%Graphs, need to redo better
\begin{figure}
  \centering
  	\includegraphics[width=1\columnwidth]{figures/silvan.pdf}
  \caption{Time series of meteorological variables for Silvan (-37.82, 145.42)}
  \label{fig:silv}
\end{figure}

@@

Combination of these variables as per equation \eqref{eq FFDI} yields the FFDI which clearly exhibits seasonality. Note the spike in summer 2009, corresponding to the Black Saturday bushfires.

%WS, Temp and FDR contours, see if can change colours to match with the diagram
\begin{figure}
  \centering
	\includegraphics[width=1\columnwidth]{figures/tw_fdr.pdf}
  \caption{Wind Speed and temperature coloured by fire danger rating}
  \label{fig:tw_fdr}
\end{figure}

The FFDI is most sensitive to temperature and windspeed \citep{dowdy10}. Figure~\ref{fig:tw_fdr} shows two-dimensional density plots for windspeed and temperature coloured by the fire danger rating (only observations from October--March included) and divided by CFA district. The relative sizes of each contour show the small proportion of extreme and code red days and the difference in weather conditions observed in different parts of Victoria. For instance the only districts observing code red days are districts 16, 17 and 18.

@@

\begin{figure}[p]
\includegraphics[width=1\columnwidth]{figures/curdist.pdf}
  \caption{Grassland curing evolution over time }
  \label{fig:cur_evol}
\end{figure}

% level plot of GS curing, probably needs resizing
\begin{figure}
  \centering
	\includegraphics[width=1\columnwidth]{figures/cur_lp.pdf}
  \caption{Level Plot of Grassland Curing }
  \label{fig:cur_lp}
\end{figure}

@@

% level plot of vapour pressure, level plot may be unnesc could just have std plot
\begin{figure}
  \centering
	\includegraphics[width=1\columnwidth]{figures/vap_ras.pdf}
  \caption{Raster plot of vapour pressure 9am }
  \label{fig:vap_ras}
\end{figure}

@@

%Commented out until final printing,
%Map of fire ignitions in Victoria
\begin{figure}
  \centering
	\includegraphics[width=1\columnwidth]{figures/firemap.pdf}
  \caption{Map of fire ignition 2000 - 2009}
  \label{fig:firemap}
\end{figure}

%timeseries histogram with count of ignitions per fortnight
\begin{figure}
  \centering
	\includegraphics[width=1\columnwidth]{figures/occurence.pdf}
  \caption{Number of ignitions by fortnight}
  \label{fig:occ}
\end{figure}

@@

\begin{figure}
  \centering
	\includegraphics[width=1\columnwidth]{figures/tw_fire.pdf}
  \caption{Wind Speed and temperature coloured by fire/no fire}
  \label{fig:tw_fire}
\end{figure}

@@

\begin{figure}
  \centering
	\includegraphics[width=1\columnwidth]{figures/dh_fire.pdf}
  \caption{Drought factor and relative humidity coloured by fire/no fire}
  \label{fig:dh_fire}
\end{figure}

@@

\begin{figure}
  \centering
	\includegraphics[width=1\columnwidth]{figures/gv_fire.pdf}
  \caption{Grassland curing and vapour pressure coloured by fire/no fire}
  \label{fig:gv_fire}
\end{figure}

@@

%eg confusion matrix with ABCD
\begin{table}
  \centering
  \begin{tabular}{lrrrr}
    & 	\multicolumn{1}{l}{} & \multicolumn{2}{c}{\emph{Model Prediction}} \\
    & 	\multicolumn{1}{l}{} & \multicolumn{1}{r}{No Fire} & \multicolumn{1}{r}{Fire} \\
    \cline{3-5}
    \emph{Actual} & No Fire & $a$   & $b$   & $a+b$ \\
                  &   Fire  & $c$   & $d$   & $c + d$ \\
    \cline{3-5}
                  &         & $a+c$ & $b+d$ & $a+b+c+d$ \\
    \cline{3-5}
  \end{tabular}
  \caption{Example confusion matrix}
  \label{table:ex}
\end{table}
An example confusion matrix is given in Table~\ref{table:ex}.
From this matrix, we can compute accuracy, sensitivity (correctly classified fires dividied by total number of actual fires), specificity (correctly identified non-fires divided by total number of non-fires), balanced accuracy and kappa:
\begin{align}
  \label{eq acc}
  \mbox{Accuracy}          & = \frac{a + d}{a+b+c+d}                     \\
  \label{eq sens}
  \mbox{Sensitivity}       & = \frac{a }{a+b}                            \\
  \label{eq spec}
  \mbox{Specificity}       & = \frac{d}{c+d}                             \\
  \label{eq BA}
  \mbox{Balanced Accuracy} & = \frac{\frac{a }{a+b} + \frac{d}{c+d} }{2} \\
  \label{eq:K}
  \kappa                   & = \frac{P_0 - P_c}{1-P_c}
\end{align}
where $P_0$ is the total agreement probability and $P_c$ is the Agreement probability which is due to chance \citep[see][for definitions]{kappa}.

On top of the accuracy measures in equations \eqref{eq sens}, \eqref{eq spec} and \eqref{eq BA} the value in the bottom left of the confusion matrix (c) where we incorrectly predict `No fire' is very important in this context.

@@

I add splines to Model 3 in stages using the confusion matrices for 10-fold cross validation to assess accuracy \citep[][using the caret package by]{caret}. The best performing model based on balanced accuracy and number of false negatives (incorrectly predicted `No Fires') is reported in Table~\ref{table:logf}.

@@

Figure~\ref{fig:gam_pl} shows the spline functions for each of the inputs. In logistic models the spline functions are not the marginal effects of the variables on probability of ignition but the marginal effect on the log-odds ratio. Equation~\eqref{eq logits} can be rearranged to give the log-odds ratio \citep{james13},
\begin{equation}
  \label{eq:logodds}
  \log(\frac{p(X)}{1-p(X)}) = \sum_{k=1}^{K} f_k (x_k).
\end{equation}

@@

%ROC for logits
\begin{figure}
  \centering
	\includegraphics[width=1\columnwidth]{figures/all_roc.pdf}
  \caption{ROC curve for logistic models}
  \label{fig:ffdi}
\end{figure}

@@

As classification trees can be sensitive to the sample used, I constructed several trees on different balanced samples of the data. As each of the trees were structured very similarly I only report two in Figure~\ref{fig:tree2}.

% Trees with all variables
\begin{figure}
  \centering
	\includegraphics[page=1, width=1\columnwidth]{figures/trees2.pdf}
	\includegraphics[page=2, width=1\columnwidth]{figures/trees2.pdf}
  \caption{Trees with all variables}
  \label{fig:tree2}
\end{figure}

@@

Interpreting the first tree in Figure~\ref{fig:tree2}, when the grassland curing level is outside the 39 to 97 range the tree predicts there will be no fire. Approximately 33\% of the training data lies in these branches.

@@

The classification tree results reaffirm the results of the spline models in Figure~\ref{fig:gam_pl} which suggested the effect of curing and drought factor on probability of ignition is not constant at each level of the variable. The bands of values chosen by the tree as having higher probability of ignition vaguely match the shape of the fitted spline functions in Model 4. The tree selecting ranges of variables for which probability of ignition changes suggests there is a  non-linear relationship between some of the explanatory variables and probability of ignition.

@@

\begin{figure}
  \centering
	\includegraphics[width=1\columnwidth]{figures/trees_roc.pdf}
  \caption{ROC curve for Tree Models}
  \label{fig:troc}
\end{figure}

@@

The variable importance is calculated as mean decrease in accuracy using the out-of-bag observations, described in detail by \citet{archer08}. Figure~\ref{fig:vimp} shows the variable importance for the forest.

@@

Similar variable importance may indicate that on the average across all trees the variables are equally important in determining probability of ignition, however variable importance cannot necessarily be taken simply at face value. \citet{strob07}  and \citet{strob08} show that variable importance measures may be biased by the measurement scale of the variables and show a preference for correlated variables. Alternative calculation of variable importance is outside the scope of this paper but the limitations of the Gini method should be taken into account.

@@

\begin{figure}
  \centering
  	\includegraphics[width=1\columnwidth]{figures/varimp.pdf}
  \caption{Variable Importance for random forest}
  \label{fig:vimp}
\end{figure}

Finally Figure~\ref{fig:rocrf} shows the ROC curve for the random forest, a single classification tree, the linear logistic model and the splines model. Table~\ref{table:AUC} contains the area under the curves (AUC) for the ROC curves shown in Figure~\ref{fig:rocrf} . We see the forest and splines model outperform a single tree and linear-logistic model. The closeness of the two models is interesting given random forests are generally expected to be a the best predictors at the expense of interpretability.

@@

\begin{figure}
  \centering
	\includegraphics[width=1\columnwidth]{figures/roc_rf.pdf}
  \caption{Variable Importance for random forest}
  \label{fig:rocrf}
\end{figure}
